# Mapping in Posit

Mapping in Posit is primarily supported by the `{sf}` package, which allows geospatial information to be held and interpreted by R functions. Then we can produce static maps using `{ggplot2}`, or interactive maps using `{leaflet}`. This guide will give details on how to construct a basic map in leaflet, but this package is very powerful, and can do a lot more. 

## Required Packages

In order to produce maps, you will need the following packages:

- `{sf}`
- `{leaflet}`
- `{sp}`
- `{raster}`
- `{terra}` 

For the purposes of mapping, the last three packages are used to support `{leaflet}` and `{sf}` (although they are spatial packages in their own right). Usually you do not need to load these directly. 

Full details on how to install these packages initially can be found on the page [How to Install and Use Geospatial R Packages](https://public-health-scotland.github.io/knowledge-base/docs/Posit%20Infrastructure?doc=How%20to%20Install%20and%20Use%20Geospatial%20R%20Packages.md) on the [Data Science Knowledge Base](https://public-health-scotland.github.io/knowledge-base/). We would recommend installing these packages now, so that you don't have to worry about getting set up when you need to produce a map.

## Setting up Posit

Any time you are doing work with any of these spatial packages, in a new session, you will need to run code to set the environment variables and load the geospatial libraries. You can do this using the code below (which is a subset of the code used to install the spatial packages!) 

<details>
  <summary>Show Initialisation Code</summary>
  
```{r eval = FALSE}
# Set environment variables to point to installations of geospatial libraries ----

## Amend 'LD_LIBRARY_PATH' ----

# Get the existing value of 'LD_LIBRARY_PATH'
old_ld_path <- Sys.getenv("LD_LIBRARY_PATH") 

# Append paths to GDAL and PROJ to 'LD_LIBRARY_PATH'
Sys.setenv(LD_LIBRARY_PATH = paste(old_ld_path,
                                   "/usr/gdal34/lib",
                                   "/usr/proj81/lib",
                                   sep = ":"))

rm(old_ld_path)

## Specify additional proj path in which pkg-config should look for .pc files ----

Sys.setenv("PKG_CONFIG_PATH" = "/usr/proj81/lib/pkgconfig")

## Specify the path to GDAL data ----

Sys.setenv("GDAL_DATA" = "/usr/gdal34/share/gdal")

# Load geospatial libraries
dyn.load("/usr/gdal34/lib/libgdal.so")
dyn.load("/usr/geos310/lib64/libgeos_c.so", local = FALSE)

```

</details>

 *If you don't run this code prior to loading spatial packages, your Posit session will abort.*

This code should be entered into your `.Rprofile` file on Posit Workbench, which means that it will automatically run when you open a session. You can edit your user wide `.Rprofile` file, which will apply to all of your Posit sessions by running the following:

```{r eval = FALSE}
# install the {usethis} package if you haven't already
# install.packages("usethis")

usethis::edit_r_profile()
```

This will open a script window where you can copy and paste the set up code. Note that if you use projects and set up a `.Rprofile` for your project specifically, you will also need to include the code in that file if you are doing something spatial.


