# Mapping in Posit

Mapping in Posit is primarily supported by the `{sf}` package, which allows geospatial information to be held and interpreted by R functions. Then we can produce static maps using `{ggplot2}`, or interactive maps using `{leaflet}`. This guide will give details on how to construct a basic map in leaflet, but this package is very powerful, and can do a lot more. 

## Required Packages

In order to produce maps, you will need the following packages:

- `{sf}`
- `{leaflet}`
- `{sp}`
- `{raster}`
- `{terra}` 

For the purposes of mapping, the last three packages are used to support `{leaflet}` and `{sf}` (although they are spatial packages in their own right). Usually you do not need to load these directly. 

Full details on how to install these packages initially can be found on the page [How to Install and Use Geospatial R Packages](https://public-health-scotland.github.io/knowledge-base/docs/Posit%20Infrastructure?doc=How%20to%20Install%20and%20Use%20Geospatial%20R%20Packages.md) on the [Data Science Knowledge Base](https://public-health-scotland.github.io/knowledge-base/).

The full script that is described on this page can be found below

<details>
  <summary>Show Installation Code - This might be a mistake to include and happy to take opinion</summary>
```{r eval = FALSE}
  # Set environment variables to point to installations of geospatial libraries ----

## Amend 'LD_LIBRARY_PATH' ----

# Get the existing value of 'LD_LIBRARY_PATH'
old_ld_path <- Sys.getenv("LD_LIBRARY_PATH") 

# Append paths to GDAL and PROJ to 'LD_LIBRARY_PATH'
Sys.setenv(LD_LIBRARY_PATH = paste(old_ld_path,
                                   "/usr/gdal34/lib",
                                   "/usr/proj81/lib",
                                   sep = ":"))

rm(old_ld_path)

## Specify additional proj path in which pkg-config should look for .pc files ----

Sys.setenv("PKG_CONFIG_PATH" = "/usr/proj81/lib/pkgconfig")

## Specify the path to GDAL data ----

Sys.setenv("GDAL_DATA" = "/usr/gdal34/share/gdal")

# Step 1 - remove geospatial R packages and their dependencies (if installed) ----

# List of geospatial packages that will be installed
geo_pkgs <- c("leaflet", "raster", "sp", "terra", "sf")

# List of geospatial package dependencies
geo_deps <- unique(
  unlist(tools::package_dependencies(packages = geo_pkgs,
                                     recursive = TRUE)))

# Remove geospatial packages and their dependencies
pkgs_to_remove <- unique(unlist(c(geo_pkgs, geo_deps)))
remove.packages(pkgs_to_remove)

# Step 2 - Install the {parallelly} package to identify number of CPUs available ----

# Remove 'parallelly' if it is already installed
remove.packages("parallelly")

# Install the 'parallelly' package
install.packages("parallelly")

# Identify the number of CPUs available
ncpus <- as.numeric(parallelly::availableCores())

# Step 3 - Install geospatial package dependencies that can be installed as binaries ----

# Get list of geospatial package dependencies that can be installed as binaries
geo_deps_bin <- sort(setdiff(geo_deps, geo_pkgs))

# Remove packages that are already installed from the list of geospatial package dependencies
geo_deps_bin <- sort(setdiff(geo_deps_bin, as.data.frame(installed.packages())$Package))

# Install these as binaries
install.packages(pkgs = geo_deps_bin,
                 repos = c("https://ppm.publichealthscotland.org/all-r/__linux__/centos7/latest"),
                 Ncpus = ncpus)

# Step 4 - Install geospatial packages from source ----
geo_config_args <- c("--with-gdal-config=/usr/gdal34/bin/gdal-config",
                     "--with-proj-include=/usr/proj81/include",
                     "--with-proj-lib=/usr/proj81/lib",
                     "--with-geos-config=/usr/geos310/bin/geos-config")
                     
# Install the {sf} package
install.packages("sf",
                 configure.args = geo_config_args,
                 INSTALL_opts = "--no-test-load",
                 repos = c("https://ppm.publichealthscotland.org/all-r/latest"),
                 Ncpus = ncpus)

# Install the {terra} package
install.packages("https://ppm.publichealthscotland.org/all-r/latest/src/contrib/Archive/terra/terra_1.7-29.tar.gz",
                 repos = NULL,
                 type = "source",
                 configure.args = geo_config_args,
                 INSTALL_opts = "--no-test-load",
                 Ncpus = ncpus)

# Install the {sp} package
install.packages("sp",
                 configure.args = geo_config_args,
                 INSTALL_opts = "--no-test-load",
                 repos = c("https://ppm.publichealthscotland.org/all-r/latest"),
                 Ncpus = ncpus)

# Install the {raster} package
install.packages("https://ppm.publichealthscotland.org/all-r/latest/src/contrib/Archive/raster/raster_2.5-8.tar.gz",
                 repos = NULL,
                 type = "source",
                 configure.args = geo_config_args,
                 INSTALL_opts = "--no-test-load",
                 Ncpus = ncpus)

# Install the {leaflet} package
install.packages("leaflet",
                 repos = c("https://ppm.publichealthscotland.org/all-r/__linux__/centos7/latest"),
                 type = "source",
                 dependencies = FALSE,
                 configure.args = geo_config_args,
                 INSTALL_opts = "--no-test-load",
                 Ncpus = ncpus)
                 

# Step 5 - Load packages ----
dyn.load("/usr/gdal34/lib/libgdal.so")
dyn.load("/usr/geos310/lib64/libgeos_c.so", local = FALSE)

library(sf)
library(leaflet)
```
</details>
<br>

## Setting up Posit

Any time you are doing work with any of these spatial packages, in a new session, you will need to run code to set the environment variables and load the geospatial libraries. You can do this using the code below (which is a subset of the code used to install the spatial packages!) 

<details>
  <summary>Show Initialisation Code</summary>
  
```{r eval = FALSE}
# Set environment variables to point to installations of geospatial libraries ----

## Amend 'LD_LIBRARY_PATH' ----

# Get the existing value of 'LD_LIBRARY_PATH'
old_ld_path <- Sys.getenv("LD_LIBRARY_PATH") 

# Append paths to GDAL and PROJ to 'LD_LIBRARY_PATH'
Sys.setenv(LD_LIBRARY_PATH = paste(old_ld_path,
                                   "/usr/gdal34/lib",
                                   "/usr/proj81/lib",
                                   sep = ":"))

rm(old_ld_path)

## Specify additional proj path in which pkg-config should look for .pc files ----

Sys.setenv("PKG_CONFIG_PATH" = "/usr/proj81/lib/pkgconfig")

## Specify the path to GDAL data ----

Sys.setenv("GDAL_DATA" = "/usr/gdal34/share/gdal")

# Load geospatial libraries
dyn.load("/usr/gdal34/lib/libgdal.so")
dyn.load("/usr/geos310/lib64/libgeos_c.so", local = FALSE)

```

</details>

This code can be saved in a network location and loaded each time using the function `source()`. Alternatively you can edit your `.Rprofile` file to include the code so that it automatically runs this code when you open a session on Posit. You can edit your user wide `.Rprofile` file, which will apply to all of your Posit sessions by running the following:

```{r eval = FALSE}
# install the {usethis} package if you haven't already
# install.packages("usethis")

usethis::edit_r_profile()
```

This will open a script window where you can copy and paste the set up code. Note that if you use projects and set up a `.Rprofile` for your project specifically, you will also need to include the code in that file if you are doing something spatial. *If you don't run this code prior to loading spatial packages, your Posit session will abort.*


